A Bayesian model of latitudinal temperature gradients applied to the Early Eocene

### 2023-02-28 manuscript/manuscript.Rmd
add references to bib

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Quantitative, geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) are used to infer palaeotemperature estimates using experimentally derived transfer functions [e.g. @Liu2009]. These data can be readily processed for regional or global-scale palaeoclimate reconstructions [@...]. 2) Palaeobiological, ecological and geological proxy data can provide broad climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Kvacek2007; @Cao2019]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Kvacek2007; @Royer2012]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply the model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Marwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990; ...]. Although some more recent climate models have been able to better align their simulations with the proxy record [@Huber2013;...], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999]. This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@hollis2019].

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32 degS), with minimum sea surface temperature (~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef]. Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO. We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N). Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.

@@

In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2008]. Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction. For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008]. The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.

@@

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Faucette2007; @Kvacek2007]. These methods have the problem that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@]. Although coral reefs are threatened by warming sea surface temperatures today [@], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).

>>>>

Existing palaeoclimate reconstructions are generally based on one of three approaches [see @Burgener2023]: 1) Geochemical proxies such as oxygen isotopes (d18O), tetraether indices (TEX86) and magnesium calcium ratios (Mg/Ca) provide quantitative palaeotemperature estimates using transfer functions based on experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. 2) Palaeobiological, ecological and lithological proxy data  provide climate constraints, but no experimentally determined transfer functions exist to directly infer quantitative climate parameters from these records [e.g. @Wolfe1995; @Kvavcek2007; @Cao2019]. Instead, palaeoclimate signals are inferred based on the climatic conditions of analogous, modern sediments, or based on the premise that the climatic requirements of ancient taxa, plant traits or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012]. 3) Earth system models offer the most highly resolved and complete climate reconstructions, but ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

@@

Here, we present a Bayesian, hierarchical model for inferring the latitudinal gradient of sea surface temperatures (SSTs) that expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern SST distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

The early Eocene is the focus of a substantial body of palaeoclimate literature [@Berggren1998; @Thomas2000; @Hyland2017; @Lunt2021], as it is the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and potentially represents an analogue for extreme climate warming scenarios [@Burke2018]. For palaeoclimate modelers, the early Eocene presents a challenge: Geochemical proxy data [@Hollis2019] and ecological data [@Markwick1994; @Greenwood1995] indicate tropical to subtropical conditions up to high latitudes, but early climate models have not been able to model warm polar temperatures and a shallow latitudinal temperature gradient under realistic conditions [@Sloan1990]. Although more recent climate models have been able to better align their simulations with the proxy record [@Huber2011], the nature of the early Eocene temperature gradient remains a matter of debate [@Tierney2017]. The shape and magnitude of the inferred EECO latitudinal gradient depends on the selection of proxies, on the method for dealing with spatial gaps and uneven sampling, on the inclusion or exclusion of terrestrial, shallow and deep marine data, while differential reporting of the results complicates comparisons of different approaches [@Evans2018; @Hollis2019; @Inglis2020]. Our goal is to provide a robust, yet simple method for inferring and comparing temperature gradients that can be readily used with different proxy types and with patchy records. We use this model to quantify the latitudinal SST gradient and the global mean SST of the early Eocene, but the unspecific design of our method means it can be readily applied to other time intervals and proxy records.

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides four different geochemical proxies for reconstructing seawater temperature: $\delta^18^O, \Delta$_47_, Mg/Ca and TEX_86_. For our analyses, this dataset was restricted to the EECO and samples from the continental shelf. Recrystallised $\delta^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@schrag1999]. This resulted in most $\delta^18^O samples being excluded from the dataset (x out of x). After data filtering, x geochemical proxy samples remained. For a detailed description of each proxy see [@Hollis2019].

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (~34°N–32 degS), with minimum sea surface temperature (~18 degC) tolerances being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained with some studies suggesting up to 35.6 degC in the geological past [@jones2022reef]. Nevertheless, the distribution of coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with some communities found up to palaeolatitudes of 43 degC N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generate quantitative sea surface temperature estimates for the ECCO. To do so, we extract localities inferred to be Ilerdian (early Eocene) coral reefs from the compilation, and that can be confidently assigned to the EECO. We exclude coral knobs and coral-bearing mounds which might have broader climatic limits than coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (<34 deg N). Subsequently, we use statistically derived temperature limits (minimum = 21 degC, average = 27.6 degC, maximum = 29.5 degC) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6 and a standard deviation of 2.125, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures > 29.5 degC, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than modern-day coral reefs.

@@

In our Bayesian, parametric model, informative priors on the model parameters improve the prediction of sea surface temperatures in the unsampled, very high latitudes: Notice that the upper limit of the credible interval does not increase beyond the range of the data, whereas unconstrained approaches such as splines, Gaussian processes or even standard linear regression could lead to unrealistically high upper bounds in this case [see @Rasmussen2004]. Prior information on the shape of latitudinal temperature gradients on Earth exists for all geological time periods, and should be used in any palaeoclimate reconstruction. For example, the greater amount of solar radiation per unit area in low latitudes causes Earth's latitudinal temperature gradient to be broadly negative [@Beer2008]. The ease with which such prior information can be integrated is a major advantage of our method, as the shape of the modeled gradient is controlled by four parameters which clearly relate to its magnitude, steepness and the latitude of its greatest steepness.

@@

Our approach for deriving fully quantitative climate reconstructions from ecological data is borrowed from nearest living relative methods, commonly employed in terrestrial, Cenozoic climate reconstructions [@Fauquette2007; @Pross2012]. These methods have the problem that the thermal preferences of taxa may have changed over time. More significantly, in the early Eocene, sea surface temperatures may have reached heights unknown in the modern world, and nearest living relative methods based on the modern are inherently unable to predict such elevated temperatures. This is especially true for taxa that inhabit the warmest part of the ocean today, e.g. coral reefs [@]. Although coral reefs are threatened by warming sea surface temperatures today [@], it is conceivable that Eocene reef corals were adapted to a warmer climate. The fossil record indicates that reef development may have been stunted in the early Eocene, but the few early Eocene coral reefs occur in low latitudes [@Zamagni2012]. Tropical temperatures predicted by the geochemical proxy record indicate hotter-than-modern tropical temperatures (Fig. S2), suggesting that the modern climate range of coral reefs may underestimate the early Eocene thermal niche for coral reefs. We have tried to account for that possibility by widening the temperature probability distribution for reef corals, but the predicted temperatures for the reef and mangrove sites still lie below the temperatures indicated by the geochemical proxy record (Fig. 4, Fig. S2).
